---
title: "FusioMRdev Tutorial"
author: "Bowei Kang, Sihao Feng, Lin S. Chen"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{FusioMRdev Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  prompt = FALSE,
  warning = FALSE
)
```

# Introduction

This vignette walks through the usage of the **FusioMRdev** package.
FusioMR is a Bayesian Mendelian randomization (MR) framework for estimating
the causal effect of an exposure on one or more outcomes from GWAS summary
statistics. It is designed primarily for molecular trait exposures (e.g.,
gene expression),where the number of available cis-QTLs as instrumental variables (IVs) is often limited and horizontal pleiotropy is pervasive. FusioMR is also
applicable to complex trait exposures with a moderate-to-large number of
IVs. For methodological details, please refer to https://doi.org/10.1016/j.ajhg.2026.03.017.

FusioMRdev provides four models, accessed via the `model` argument of
`fusiomr()`:

| Model            | Exposure | Outcome | Use when                                              |
|------------------|----------|---------|-------------------------------------------------------|
| `seso_uhp_only`  | 1        | 1       | uncorrelated pleiotropy (UHP) is the concern |
| `seso_with_chp`  | 1        | 1       | correlated pleiotropy (CHP) is the concern   |
| `semo`           | 1        | 2       | one exposure, two outcomes                     |
| `memo`           | 2        | 2       | two exposures, two outcomes;    

Throughout this vignette, **UHP** stands for *uncorrelated horizontal
pleiotropy* and **CHP** for *correlated horizontal pleiotropy*.

The input format mirrors that of `TwoSampleMR` and `MendelianRandomization`:
the user supplies SNP–exposure and SNP–outcome effect estimates and their
standard errors. FusioMRdev assumes that the input SNPs have already been
**clumped to approximate independence** and **selected as IVs** (e.g., by a
p-value threshold on the exposure). The package does not perform clumping
or IV selection internally.

# Installation

FusioMRdev requires **R >= 4.3.0** and a working C++ compiler:

- **macOS**: `xcode-select --install`
- **Windows**: install [Rtools](https://cran.r-project.org/bin/windows/Rtools/)

Install the development version from GitHub:

```{r install, eval = FALSE}
# install.packages("devtools")
devtools::install_github("fsh56/FusioMR-software")
```

Load the package:

```{r load}
suppressMessages(library(FusioMRdev))
```

The packages `Rcpp`, `RcppArmadillo`, and `invgamma` are pulled in
automatically as dependencies.

# Single exposure, single outcome: `seso_uhp_only`

The `seso_uhp_only` model handles the most common MR setup: one exposure,
one outcome, and uncorrelated horizontal pleiotropy as the main concern.
It is the recommended default for TWMR analysis with limited IVs.

We use a pre-generated example dataset, simulated under the limited-IV
TWMR setting (exposure GWAS sample size = 300, outcome GWAS sample size =
20,000, true causal effect $\beta = 0$, IVs selected at p < 1e-3).

```{r seso-uhp-load}
d1 <- readRDS("~/examples/data/seso_uhp_only_example.rds")
str(d1, max.level = 1)
```

The four required inputs are vectors of equal length: SNP–exposure
effects (`b_exp`), their standard errors (`se_exp`), and the analogous
quantities for the outcome (`b_out`, `se_out`).

Run the model:

```{r seso-uhp-fit}
fit1 <- fusiomr(d1$b_exp, d1$se_exp, d1$b_out, d1$se_out,
                model = "seso_uhp_only")
```

The returned object is a list with the posterior summary of the causal
effect:

```{r seso-uhp-results}
fit1$est    # posterior mean of beta
fit1$se     # posterior SD
fit1$pval   # Bayesian tail probability (pseudo p-value)
fit1$ci     # 95% empirical credible interval
```

# Single exposure, single outcome with CHP: `seso_with_chp`

When some IVs are suspected of having correlated horizontal pleiotropy
(for example, when the exposure is a complex trait with many IVs and a
subset may share confounders with the outcome), use `seso_with_chp`. The
model adds a latent CHP indicator on each IV; label-switching is
corrected internally.

We use a complex-trait example (exposure and outcome GWAS sample sizes
both 20,000, true $\beta = 0$, 5% of IVs have CHP, IVs selected at
p < 1e-5).

```{r seso-chp-fit}
d2 <- readRDS("~/examples/data/seso_with_chp_example.rds")

fit2 <- fusiomr(d2$b_exp, d2$se_exp, d2$b_out, d2$se_out,
                model = "seso_with_chp")

fit2$est
fit2$se
fit2$pval
fit2$ci
```

# Single exposure, two outcomes: `semo`

`semo` jointly analyzes one exposure and two correlated outcomes. The
model places a bivariate prior on the per-IV UHP effects across the two
outcomes, allowing information sharing while estimating each
exposure–outcome causal effect independently.

The input format differs from the single-outcome models: `b_out` and
`se_out` must be `K x 2` matrices, where column $j$ contains summary
statistics for outcome $j$.

We use a TWMR-style example with one shared exposure and two outcomes
(exposure sample size = 300, outcome 1 sample size = 80,000, outcome 2
sample size = 20,000, true causal effects $\beta_1 = \beta_2 = 0$, UHP
correlation 0.4).

```{r semo-fit}
d3 <- readRDS("~/examples/data/semo_example.rds")
dim(d3$b_out) 

fit3 <- fusiomr(d3$b_exp, d3$se_exp, d3$b_out, d3$se_out,
                model = "semo")

fit3$est
fit3$se
fit3$pval
fit3$ci
```

# Two exposures, two outcomes: `memo`

`memo` is the most general model in FusioMRdev. It jointly analyzes two
correlated exposures and two correlated outcomes, accounting for both
UHP and CHP. The summary-statistics inputs become `K x 2` matrices for
both the exposure side and the outcome side: column $j$ corresponds to
exposure $j$ (or outcome $j$).

The call signature is:

```{r memo-call, eval = FALSE}
# b_exp, se_exp, b_out, se_out are all K x 2 matrices
fit4 <- fusiomr(b_exp, se_exp, b_out, se_out, model = "memo")

fit4$est    # length-2 vector
fit4$se
fit4$pval
fit4$ci     # 2-row matrix
```

`memo` is more computationally demanding than the other three models
because it samples two bivariate covariance matrices ($\Sigma_\gamma$
and $\Sigma_\theta$) at each Gibbs iteration.

# Tuning: `parameter_control()`

For most users, the defaults are sufficient and `parameter_control()`
does not need to be called explicitly. When tuning is needed, pass a
customized control list to `fusiomr()`:

```{r control-example, eval = FALSE}
ctrl <- parameter_control(
  niter = 30000,                   # longer MCMC
  z_thresh = qnorm(1 - 5e-8 / 2),  # winner's-curse correction
  rho_ov  = 0.2                    # sample overlap correction
)

fit <- fusiomr(b_exp, se_exp, b_out, se_out,
               model = "seso_uhp_only", control = ctrl)
```

A summary of the available tuning parameters:

| Argument             | Purpose                                                                                    |
|----------------------|--------------------------------------------------------------------------------------------|
| `niter`              | Number of Gibbs iterations (default 20000).                                                |
| `burnin_prop`        | Burn-in proportion in [0, 1) (default 0.5).                                                |
| `c_gamma`, `c_theta` | Prior weight per IV for $\sigma^2_\gamma$ and $\sigma^2_\theta$. Larger values make the prior more concentrated around its mean. |
| `kappa_gamma`, `kappa_theta` | Conservative inflation multipliers (>= 1) for the prior means.                     |
| `Kmin`, `Kmax`       | Floor and cap on $K$ when computing prior shape parameters.                                |
| `rho_ov`             | Sampling correlation between exposure and outcome due to sample overlap, in [-1, 1].       |
| `z_thresh`           | Optional $|Z_\gamma|$ selection threshold (winner's-curse correction). For p = 5e-8 two-sided, use `qnorm(1 - 5e-8/2)`. |
| `trim`               | Tail probability for winsorized moments (default 0.1).                                     |
| `hybrid`             | Logical; enable hybrid empirical-Bayes priors (see next section).                          |
| `kappa_hybrid`       | Pooling strength for hybrid mode; larger values shrink the local moment toward the global. |
| `global_mean_gamma`, `global_mean_theta` | Global EB centers (scalars), required for hybrid mode in `seso_*` models. |
| `global_Sigma_gamma`, `global_Sigma_theta` | Global EB centers (2x2 matrices), required for hybrid mode in `semo` / `memo`. |

See `?parameter_control` for full details.

# Hybrid empirical-Bayes priors (`hybrid = TRUE`)

By default, FusioMR sets the variance priors using a method-of-moments
estimate computed from the input IVs alone (the *local* moment).
In gene-region-specific TWMR analyses where the number of IVs per region
is small, the local moment can be noisy. The hybrid mode pools the local
moment with a global empirical-Bayes center estimated from a larger
reference set.

Setting `hybrid = TRUE` requires the user to supply the appropriate
global center(s) for the chosen model:

| Model           | Required global argument(s)                           | Type            |
|-----------------|-------------------------------------------------------|-----------------|
| `seso_uhp_only` | `global_mean_gamma`, `global_mean_theta`              | scalar, scalar  |
| `seso_with_chp` | `global_mean_gamma`, `global_mean_theta`              | scalar, scalar  |
| `semo`          | `global_mean_gamma`, `global_Sigma_theta`             | scalar, 2x2 mat |
| `memo`          | `global_Sigma_gamma`, `global_Sigma_theta`            | 2x2 mat, 2x2 mat |

The typical workflow is two-pass:

1. **Estimate the global centers** by running FusioMR on a large
   reference set of exposures (e.g., all genes in a TWMR scan) and
   summarizing the per-exposure local moments (e.g., taking the median
   across exposures).
2. **Refit the target exposure** with `hybrid = TRUE`, passing the
   global centers to `parameter_control()`.

## pseudo-code: `seso_uhp_only` with hybrid priors

```{r hybrid-seso, eval = FALSE}
# After computing global_mean_gamma and global_mean_theta from a reference scan:

ctrl_hybrid <- parameter_control(
  hybrid            = TRUE,
  kappa_hybrid      = 5,
  global_mean_gamma = global_mean_gamma,   # your global estimate (scalar)
  global_mean_theta = global_mean_theta    # your global estimate (scalar)
)

fit <- fusiomr(b_exp, se_exp, b_out, se_out,
               model = "seso_uhp_only", control = ctrl_hybrid)
```

## pseudo-code: `semo` with hybrid priors

```{r hybrid-semo, eval = FALSE}
ctrl_hybrid <- parameter_control(
  hybrid             = TRUE,
  kappa_hybrid       = 5,
  global_mean_gamma  = global_mean_gamma,   # your global estimate (scalar)
  global_Sigma_theta = global_Sigma_theta   # your global estimate (2 x 2 matrix)
)

fit <- fusiomr(b_exp, se_exp, b_out, se_out,
               model = "semo", control = ctrl_hybrid)
```

## Pseudo-code: `memo` with hybrid priors

```{r hybrid-memo, eval = FALSE}
ctrl_hybrid <- parameter_control(
  hybrid             = TRUE,
  kappa_hybrid       = 5,
  global_Sigma_gamma = global_Sigma_gamma,  # your global estimate (2 x 2 matrix)
  global_Sigma_theta = global_Sigma_theta   # your global estimate (2 x 2 matrix)
)

fit <- fusiomr(b_exp, se_exp, b_out, se_out,
               model = "memo", control = ctrl_hybrid)
```

If `hybrid = TRUE` is set but the required global argument(s) are
missing or have the wrong shape, `fusiomr()` stops with an informative
error message before running the sampler.

# Getting help

```{r help, eval = FALSE}
?fusiomr
?parameter_control
```

Bug reports and feature requests:
[github.com/fsh56/FusioMR-software/issues](https://github.com/fsh56/FusioMR-software/issues).

# Session info

```{r session}
sessionInfo()
```
